---
title: "Subsetting"
lang: "en"
output:
  html_document:
    theme: paper
    highlight: textmate
    toc: true
    toc_depth: 5
    toc_float: true
---

```{r setup, include=FALSE}
source("ma_tings.R") 
```
Sub-setting allows us to select specific elements within a **vector** of items.

#### Selecting elements by position

We can select elements by position (or **index**) in the vector.

- `x[4]`{.R} Select the fourth element

- `x[-4]`{.R} Select everything **but** the fourth element

- `x[2:5]`{.R} Select element two to five

- `x[-(2:5)]`{.R} Select everything **but** element two to five

- `x[c(1, 3)]`{.R} Select element one and three

**Trying it out!**

Let us see how some of these operations will look in [R]{.RR}. First, we'll create a vector with some values in it, and then we'll perform some operations on it. 

```{r}
x <- floor(runif(n = 10, min = 1, max = 20)) #  create a vector filled with random integers
```

```{r}
x # print out the values to terminal
```

```{r}
x[4] # select the fourth element
```

```{r}
x[-4] # select everything but the fourth element
```

```{r}
x[2:5] # select element two to five
```

```{r}
x[-(2:5)] # select everything **but** element two to five
```

```{r}
x[c(1, 3)] #elect element one and three
```

#### Logical operations (basic)

To select elements that have a specific value, we'll first ask [R]{.RR} to tell us the **index** of those items. To do this we'll just use one of the **logical operators**. Some examples of **logical operators** are:

  - `==`{.R} for equal to

- `!=`{.R} for **not** equal to

- `>`{.R} for greater than

- `<`{.R} for less than

- `<=`{.R} for less than or equal to

- `>=`{.R} for greater than or equal to

To use these logical operators we need our vector, the operator, and the value to compare it to.

For example, `x >= 1`{.R} means "which elements in `x`{.R} are greater than or equal to 1?"

When we perform logical operations on a vector like this [R]{.RR} will tell us which elements match the logical rule and which don't. It'll print out `TRUE`{.R} for those that match and `FALSE`{.R} for those that don't.
[]{#example}

**Trying it out!**

Let us see what some of these operations will look like in [R]{.RR}. We'll again create a vector with some values in it, and then we'll perform some operations on it.

```{r}
x <- c(2, 8, 11, 10, -1) #  create a vector  with some numbers

x # print out the values to terminal
```

```{r}
x >= 1 # which elements are greater than or equal to 1
```

```{r}
x == 2 # which elements are equal to 2
```

```{r}
x != 2 # which elements are NOT equal to 2
```

We can also use logical operators to find which elements in a vector are a **member of a set**.

- `%in%`{.R} for *is an member of a set*

To use the `%in%`{.R} we'll also need another vector (our comparison set)

**Trying it out!**

```{r}
x # print out x just is case we forgot what was in it!
```

```{r}
x %in% c(1, 8, -1) # elements a member of the set {1, 8, -1}
```

Once we know which elements of a vector match a logical rule there are a couple of things we can do with this information: 1) we can ask for the element **indexes** or 2) we can ask for the element **values**

##### Getting element indexes

To get to positions in a vector that match a particular rule we just wrap our logical operation in a `which()`{.R} function.

For example, `which(x > 5)`{.R} asks "what are the positions in `x`{.R} that have a value greater than 5?"

**Trying it out!**

```{r}
x # print out x just is case we forgot what was in it!
```

```{r}
which(x == 11) # which position is equal to 11
```

```{r}
# the next line will return an empty vector because no values are greater than 15
which(x > 15) # which position is greater than 15
```

```{r}
# we can save the output of our logical operation to a variable and then use that as input for the which() function

matches <- x == 11
which(matches)
```

##### Getting element values

We saw that when we used a logical operator on a vector is returned a vector of `TRUE`{.R} and `FALSE`{.R} values. We can use these `TRUE`{.R} and `FALSE`{.R} values to select elements in vector.

- `x[x > 0]`{.R} Select all element that have a value greater than 0

- `x[x %in% c(1, 6, 8)]`{.R} Select all elements that are a member of the set {1, 6, 8}

- `x[-(2:5)]`{.R} Select everything **but** element two to four


**Trying it out!**

Let us see how some of these operations will look in [R]{.RR}. We'll again create a vector with some values in it, and then we'll perform some operations on it.

```{r}
x <- c(10, 10, 10, 11, -1) #  create a vector  with some numbers
```
```{r}
x # print out the values to terminal
```
```{r}
x[x == 10] # all the elements with value 10
```
```{r}
matches <- x == 10

x[matches] # note that there's no '' around matches because it's a variable name
```

#### Selecting elements from named vector

Finally, when we have **named** vectors we can select elements using the **name** of that element

- `x["arms"]`{.R} Select the element named **arms**

**Trying it out!**

```{r}
x <- c(arms = 2, legs = 2, eyes = 2, heads = 1)
x
```

```{r}
x['arms']
```



##### Appendix



<details><summary>Show code</summary>
```{r eval=FALSE, code=xfun::read_utf8("subsetting.r")}
```
</details>

[ss]{#example}

```{r copy=FALSE,echo=FALSE}
require(formatR)

junipr::mywidget(paste0(xfun::read_utf8("subsetting.r")[2:55],collapse = "\n"), elementId = "junipr") %>% widgetframe::saveWidgetframe(file = "test")
```

